## Introduction to scripted data pre-processing in R {#module13}

This module is meant for researchers who do not yet used code scripts but who
are interested in starting or are supervising other researchers who are working
with code for biomedical analysis. Our aim in this module is to provide enough
information that someone without coding experience can gain some comfort in
navigating R code scripts, either to help understand a paper that includes such
scripts as part of its supplemental materials or to help understand the work of
a trainee who is incorporating code in their research. For researchers who are
already using code scripts, we recommend the next module, which provides some
advice on steps that can improve reproducibility when writing scripts for
biomedical data preprocessing.

In this module, we will provide an introduction to scripted pre-processing of
experimental data through R scripts. We will introduce the basic elements of an
R code script as well as the basics of creating and running a script. At the end
of the module, through a video exercise, we will demonstrate how to create,
save, and run an R code script for a simple data preprocessing task.

**Objectives.** After this module, the trainee will be able to:

- Describe what an R code script is and how it differs from interactive
coding in R 
- Explain how code scripts can increase reproducibility of data pre-processing
- Create and save an R script to perform a simple data pre-processing task 
- Run an R script
- Work through an example R script using a video exercise

### What is a code script?

The simplest method of working with R is through something called *interactive
coding*. With this style of coding, you enter a single command or function call
at the cursor in the console, tell the program to execute that one element of
code (for example, by pressing the Return key), and then wait until it executes
it before you enter the next command or function call.

A script, on the other hand, is a longer document that gives all the steps in a
process. You can think of a code script as being like a script for a play---it's
a record of everything that happens over the course of the event. For a play,
the script records the dialogue and stage directions for a play, while for a
data preprocessing task, it can record all the steps from inputting the data
through preprocessing steps and finally saving the data in a processed form for
further analysis, visualization, and statistical testing.

You can run the same code whether you're using a script or typing in the
commands one at a time in the console as interactive coding. However, when you
code interactively at the console, you're not making a record of each of your
steps (as a note, there are ways to save the history of commands typed at a
console, but it can be very messy to try to use later to reproduce and remember
what you did originally, so you should consider commands that are typed at the
console to not be recorded for the purposes of reproducibility). When you write
your code in a script, on the other hand, you have a record that you can
later reopen to see what you did or to repeat the steps. In a very broad way,
you can visualize this process as walking in wet sand---you are making a record
(footsteps) of the path you took while you are making that path.

A code script is typically written in a plain text document, and you can create,
edit, and save code scripts in any interactive development environment (like
RStudio if you are programming in R). The program (R for example) can then read
and run this script as a "batch" at any time. In other words, it can walk
through and execute each piece of code that you recorded in the script, rather
than you needing to enter each line of code one at a time in the console. For
many programming languages, you can also run the code in a script in smaller
sections, executing just one or a few lines at a time to explore what's
happening in each line of the code. With this combination of functionality, as
well as recording of code for future reference or reproduction, code scripts
provide an excellent method for building and using pipelines of code to
preprocess biomedical data.

In later sections of this module, we'll walk through the practical steps of
writing one of these code scripts. In a video exercise at the end, we'll look at
an example script for a simple task in biomedical data preprocessing,
calculating the rate of growth of bacteria under different growing conditions.
In this exercise, we'll walk you through how to open, run, and explore this
script in RStudio.

### How code scripts improve reproducibility of preprocessing

In the introduction to this book, we provide the definition for computational
reproducibility. Specifically, computational reproducibility typically means
that another researcher could get the exact results of the original study from
the original data collected from a study [@stark2018before]. Computational
reproducibility, then, requires a few things: the original data and very
thorough instructions that describe how those data were processed and analyzed
[@nekrutenko2012next].

Neither of these elements is trivial to provide in a thorough way for a complex
biomedical experiment. Raw datasets are often extremely large and complex. To
provide thorough instructions on the processing and analyis requires "access to
... source code or binaries of exact versions of software used to carry out the
initial analysis (this includes all helper scripts that are used to convert
formats, groom data, and so on) and knowing all parameter settings exactly as
they were used" [@nekrutenko2012next].

By using a code script for data preprocessing (and data analysis and
visualization), you can often substantially improve the computational
reproducibility of your experiment, because the code script itself documents the
exact and precise instructions for how the data are processed and analyzed. For
example, an R script will include all instructions for how the data were loaded
from a file, and will even include the file name where the data are saved, as it
must reference this to input the data. Further, it provides a list of all the
function calls that were run and the order in which they were run. For each
function call, it provides the details on the parameter settings used for that
function. Since R is an open-source language, and its packages are largely
open-source as well, if you know the version of R and each package used in the
script, you can find and read through all the underlying code that defines all
the functions used in the script. In other words, the open-source nature of the
code means that you can, if you want, dig into the algorithms underlying any
step of the process, and so you do not have to consider any step of the script
as a "black box".

In other words, in the course of writing an executable script to preprocess
data, you are thoroughly documenting each step that you take in that process,
creating one of the key components (clear instructions on how the data were
processed and analyzed) that is necessary to make an experiment computationally
reproducible. Once you have this script, there are only two other elements
that are required to make the experiment fully computationally reproducible: 
first, the original, raw data, and second, information on the versions of 
any software you used in the code (this would include the version of R that
was used, as well as versions of any R packages that were used to 
supplement the base R functions). 

### How to write an R code script

In this section, we'll go through some basics to help you get started writing a
code script in R. The process of writing a code script is similar in many other
interpreted languages, like Python and Julia. If you are familiar with writing
code scripts in R, find the next module---where we provide some tips on
improving reproducibility when writing scripts---more helpful.

We'll start with a few basics of the conventions of the R programming language.
If you have never used R before, it is critical to understand these basic
pieces---just enough so you can understand how an R code script is put together
and run. In later modules, we'll go into some more detail about some helpful
tools in R, including Bioconductor data structures and the suite of "tidyverse"
tools that are now taught in most beginner R programming courses. We will, of
course, not have room to provide a full course on how to program in R, but we
are aiming to give you enough of a view that you can understand how R
programming can fit in to the data preprocessing and analysis pipeline for
laboratory-based biomedical research projects, as well as how you can navigate
an R script that someone else has written. In a later module, we'll provide
directions to more resources if you would like to continue developing your
expertise in R programming beyond the basics covered in these modules.

**What is an R object?**

First, you'll need to understand where R keeps data while you're working with
it. When you work in R, any piece of data that you work with will be available
in something called an "object". The simplest way to think of this R object is
simply as a container for data. Different objects can be structured in different
ways, in terms of how they arrange the data---which has implications for how
you can access the data from that object---but regardless of this structure, all
R objects share the same purpose of storing data in a way that's available to
you as you work in R.

One of the first steps in most R scripts, therefore, will be to create some of
these objects. Until you have some data available, there's not much interesting
stuff that you can do in R. If you want to work with data that are stored in a
file---for example, data that you recorded in the laboratory and saved in an
Excel file---then you can create an R object with that data by reading in the
data using a specific R function (we'll cover these in a minute). This will read
the data in R and store it in an object where you can access it later.

To keep track of the objects you have in your R session, you typically assign
each object a name. Any time you want to use the data in that object, or work 
with the object in any way, you can then refer to it by that name, rather than 
needing to repeat all the code you used to initially create it. You can assign 
an object its name using a special function in R called the "gets arrow" or 
assignment operator. It's an arrow made of the less than and hyphen keys, with 
no spaces between the two (`<-`). You'll put the name you want to give the object
to the left of this arrow and the code to create the object (for example, to read
in data from a file) to the right. Therefore, the beginning of your R script
will often have one or more lines of code that look like this: 

```{r eval = FALSE, echo = TRUE}
my_data <- read_excel("my_recorded_data.xlsx")
```

In this example, the line of code is reading in data from an Excel file named
"my_recorded_data.xlsx" and storing in an R object that is assigned the name 
`my_data`. When you want to work with these data later in the code pipeline, you 
can do so by referencing `my_data`, which now stores the data from that file. 

In addition to creating objects from the data that you initially read in, you will 
likely create more intermediate objects along the way. For example, if you take
your initial data and filter it down to a subset, then you might assign that version 
of the data to a separate object name, so you can work with that version later in 
your code. Alternatively, in some cases you'll just overwrite the original object
with the new version, using the same object name (for example, creating a subset of
the `my_data` object and assigning it the same name of `my_data`). This reassigns the
object name---when you refer to `my_data` from that point on, it will contain the 
subsetted version. However, in some cases this can be useful because it helps keep 
the collection of R objects you have in your session a bit smaller and simpler. What's 
more, you can make these changes to simplify the version of the data you're working 
with in R without worrying about it changing your raw data. Once you read the data 
in from an outside file, like an Excel file, R will work on a copy of that data, not
the original data. You can make as many changes as you want to the data object in R
without it changing anything in your raw data. 

**What are R functions and an R function calls?**

The next key component of the R programming language is the idea of R functions
and R function calls. These are the parts of R that do things (whereas the objects in R
are the "things" that these functions operate on). An R function is a tool that can 
take one or more R objects as inputs, do something based on those inputs, and return a 
new R object as the output (occasionally they'll also have "side effects" beyond returning
this R object---for example, some functions will make a plot and show it in the plotting 
window of RStudio). 

The R objects that you input can be ones that you've assigned to a name (for
example, `my_data`). They can also be simple objects that you make on the fly,
just to have to input to that function. For example, if you're reading in data
from a file, one of the R object inputs you'll need to give the function is the
path to that file, which you could either save as an object (e.g.,
`my_data_filepath <- "my_recorded_data.xlsx"` and then reference
`my_data_filepath` when you call the function) or create as an object on the fly
when you call the function (e.g., just put `"my_recorded_data.xlsx"` directly in
the function call, as shown in the example above).

The function itself is the tool, which encapsulates the code to do something with 
input objects. When you use that tool, it's called *calling* the function. Therefore, 
all of the lines of code in your script will give *function calls*, where you are 
asking R to run a specific function (or, in some cases, a linked set of functions) 
based on specified inputs. 

For example, the following function call would read in data from the Excel file 
"my_recorded_data.xlsx": 

```{r eval = FALSE, echo = TRUE}
read_excel("my_recorded_data.xlsx")
```

This line of code is calling the function `read_excel`, which is a tool for inputting
data from an Excel file into an R object with a specific data structure. By running 
this line of code, either at the console or in an R script, you are asking R to input
data from the file named "my_recorded_data.xlsx", which is the R object that you're 
giving as an input to the function. This particular call would only read the data in---it
won't assign the resulting object to a name, but instead will just print out the data
at the R console. 

If you'd like to read the data in and save it in an object to use later, you'll
want to add another function to this call, so that you assign the output object
a name. For this, you'll use the gets arrow that we described earlier. This is a
special type of function in R. Most R functions consist of the function's name,
followed by parentheses inside of which you put the objects to input to the
function (e.g., `read_excel("my_recorded_dat.xlsx"`). The gets arrow is a
different type of function called an operator. These functions go between two
objects, both of which are input to the operator function. They're used often
for arithmetic (for example, the `+` operator adds the values in the objects
before and after it, so that you can call `1 + 2` to add one and two). For the
gets arrow, it will go between the name that you want to assign to the object
(e.g., `my_data`) and the function call that creates that object (e.g.,
`read_excel("my_recorded_data.xlsx")`):

```{r eval = FALSE, echo = TRUE}
my_data <- read_excel("my_recorded_data.xlsx")
```

In this case, the line that R will execute will include two functions, where the 
output of one gets linked straight into the second, and the result will be the 
output from the second function (that the data in the Excel file is stored in 
an object assigned the name `my_data`). 

As you write an R script, you will use function calls to work through the 
preprocessing steps. You can use different function calls to do things like
apply a transformation, average values across groups, or reduce dimensions of
a high-dimensional dataset. Once you've preprocessed the data, you can also use 
function calls to run statistical tests with the data and to visualize results
through figures and tables. 

The process of writing a script is normally very iterative---you'll write the
code to do the first few steps (e.g., read in the data), look at what you've
got, plan out some next steps, try to write some code for those steps, run it
and check your output, and so on. The process is very similar to drafting a
paper. You can try things out in early steps---and some steps won't work out at
first, or it will turn out that you don't need them. As you continue, you'll
refine the script, editing it down to the essential steps and making sure each
function call within those steps is operating as you intend. While it can be
intimidating to start with a blank file and develop some code---just like it is
with a blank piece of paper when writing a manuscript---just like with writing,
you are able to start with something rough and then iterate until you arrive at
the version you want.

This process might seem a bit overwhelming when you first learn it, but it
suffices at this point if you understand that, in R code, you'll be working with
objects (your materials) and functions (your tools). As we look through R
scripts in later parts of this module, we'll see these two pieces---objects and
functions---used again and again in the scripts. They are the building blocks
for your R scripts.

**What is an R library?**

There's one last component of R that will be helpful to understand as we move through
the rest of this module and the next few modules. That's the idea of an R package, and 
fortunately, it's a pretty straightforward one. 

We just talked about how functions in R are tools, which you can use to do interesting 
things with your data (including all the preprocessing steps we talked about in the
last module). However, the version of R that you initially install to your computer
(available for free for all major operating systems at https://cran.r-hub.io/) doesn't
include all the tools that you will likely want to use. The initial download gives you 
the base of the programming language, which is called base R, as well as a few 
extensions of this for very common tasks, like fitting some common statistical models. 

Because R is an open-source software, people who use R can build on top of this
simple base. R users can create new functions that combine more rudimentary
tools in base R to create customized tools suited to their own tasks. R users
can create these tools for their own personal use, and often do, but there is
also a mechanism for them to share these new tools with others if they'd like.
They can bundle a set of R functions they've created into an *R package* and
then post this package on a public repository where others can download it and
use the functions in it. In some of the examples in these modules, we'll be
using tools from these packages, and it's rare that someone uses R without using
at least some of these supplementary packages, so it's good to get an idea of
how to get and use them.

The people who make packages can share them in a number of repositories, but the
most standard repository for sharing R packages widely is the Comprehensive R
Archive Network (CRAN). If a package is shared through CRAN, you can get it
using the function `install.packages` along with the package's name. For
example, in the code we showed earlier, the `read_excel` function does not come
with base R, but instead is part of a package called `readxl`, which is shared
on CRAN. To download that package so that you can use its functions, you can
run:

```{r eval = FALSE, echo = TRUE}
install.packages("readxl")
```

This will download the code for the package and unpack it in a special part of
your computer where R can easily find it. You only need to install a package
once, at least until you get a new computer or update your version of base R.
However, to use the functions in that package, you'll need to *load* the package
in your current R session. This makes the functions in that package available to
you as you work in that R session. To do this, you use the `library` function,
along with the name of the package. For example, to load the `readxl` package in
an R session, you'd need to run:

```{r eval = FALSE, echo = TRUE}
library("readxl")
```

While you only need to *install* a package once, you need to *load* it every
time you open a new R session to do work, if you want to use its functions in
that R session. Therefore, you'll often see a lot of calls to the `library`
function in R scripts. You can use this call anywhere in the script as long as
you put it before code where you use the library's functions, but it's great to
get in the habit of putting all the `library` function calls at the start of
your R script. That way, if you share the script with someone else, they can
quickly check to see if they'll need to install any new packages before they can
run the code in the script.

**Using a plain text file, edited in a plain text editor (can open in RStudio IDE)**

Based on the points that we've just discussed, hopefully you can envision now
that an R script will ultimately include a number of lines of code, covering a
number of R function calls that work with data stored in objects. You can expect
there to be lots of calls that assign objects their own names (with `<-`), and
the function calls will typically include both a function called by name and
some objects as input to that function, contained inside parentheses after the
function name.

This type of script should be written in plain text, and so the best way to
create an R script is by using a text editor. Your computer likely came with a
text editor as one of the pieces of utility software that was installed by
default. However, with R scripts, it can be easier to use the text editor that
comes as part of RStudio. This allows you to open and edit your scripts in a
nice environment, one that includes a console area where you can test out pieces
of code, a pane for viewing figures, and so on.

In RStudio, you can create a new R script by going to the "File" menu at the top
of the screen, choosing "New File" and then choosing "R Script". This will open
a new plain text file that, by default, will have the file extension ".R" (e.g.,
"my_file.R"), which is the standard file extension for R scripts. Once you've
created an R script file, you can begin writing your script. In the next
section, we'll walk through how you can run code that you've put into your
script. However, we think it's worth mentioning that, as you get started on this
process, you might find it easiest to start not by writing your own R script
from scratch, but instead by starting with someone else's and walking through
that. You can explore how it works (reverse engineer it). Then you can try
changing small parts, to see if it acts as you expect when you do. This process
will help you get a feel for how these scripts are organized and how they
operate. Later in this module, we'll provide an R script for a basic laboratory
data preprocessing task and walk you through it, so you can use that as a
starting point to understand how it would work to create, edit, and run your own
R script.

### How to run code in an R script

Once you've written code in an R script, you can run (execute) that code in a
number of ways. First, you can run all the code in the script at once, which is
known as *batch execution*. When you do this, all the code in the script will be
executed by R, and while it's executed by R one line at a time, you won't have
the chance to make changes along the way. If you compare it to the idea of a
code script as like a play script, you can think of this as being like when the
play is performed for an audience---you start the play, but then you don't have
the chance to stop and work on it as it's going. Instead, it will go straight
through to the end. If there is an error somewhere along the way, then the code
will stop running at that point and you'll get an error message, but otherwise
when you run the code as a batch, R won't stop executing the lines until it gets
to the end. This mode of running the code is great for once you've developed a
pipeline that you're happy with---it quickly runs everything and provides the
output.

The other way that you can execute the code is by running a single line, or a
small set of lines, of the code at a time. In the play analogy, this is similar
to what might happen during rehearsals, when you go through part of the play
script and then stop to get comments from the director, then either re-try that
part with a few changes or move on to the next small part. This mode of running
the code is great for when you're developing the pipeline. Just like with a
play's rehearsals, you'll want a lot of chances to explore and change things as
you develop the final product, and this mode of running code is excellent for
exploration and editing. Often, most of your time when you code will be spent
doing this style of code execution. Running in batch mode will get a lot of work
done, but is very quick for the programmer---developing the code is what takes
time, and just like with writing a manuscript, this time comes from drafting a
rough draft and then editing it until you arrive at a clean and clear final
version.

Both of these methods of code execution are easy to do in RStudio. Since you'll
usually start by using line-by-line execution, we'll start with showing how you
can do that. In RStudio, you can open your code script (a file ending in ".R"),
and you will still be able to see the console, which is a space for submitting
function calls to R. To execute the code in the script one line at a time,
there's a few quick ways that you can tell RStudio to send that line in the
script to the console to run. Start by putting your cursor on that line of code.
One way to now execute this line (i.e., send it to the console to run) is to
click on the "Run" button in the top right-hand corner of the script file. If
you try this, you should see that this line of code gets sent to the console
pane of RStudio, and the results from running that line are shown in the
console.

Even quicker is a keyboard shortcut that does the same thing. (Keyboard
shortcuts are short control sequences that you type in your keyboard to run a
command. They're faster than clicking buttons because you can do them without
taking your hands off the keyboard. Ctrl-C is one very common one that you might
have used before, which in most programs will copy the current selection.) With your
cursor on the line of the function call that you want to execute, use the keyboard
shortcut Ctrl-Enter (depending on your operating system, you may need to use
Command rather than Ctrl).

You can use a similar method to run a few lines of code at once. All you have to 
do is highlight the code that you want to run, and then you can use either of 
the two methods (click the "Run" button or use the Ctrl-Enter keyboard shortcut). 
We will show you examples of how to do this in the video exercise at the end
of this module.

To execute an R script in batch mode, there are again a could of ways you can do
it. First, there is a "Source" button in the top right of the R script file when
you open it in RStudio. You can click on this button and it will run the entire
script as a batch. There is also an R command that you can use to source a file
based on its file name, `source`. If you have a file in your working directory
named "my_pipeline.R", for example, you can execute the code in it in a batch by
running `source("my_pipeline.R)".

To get started, it's probably easiest to just use the buttons "Run" and "Source" 
that RStudio provides in the window for the R script file. As you do more work, you
may find some of these other methods help you work faster, or allow you to do
more interesting things, so it's good to know they're there, but you don't need
to try to navigate them all as you learn how to run code in an R script. 

### Exercise

As we mentioned earlier, it can be helpful as you learn how to navigate R
scripts to start by dissecting and exploring an existing script. In the video
that is embedded here, we give an example of that using a script that captures
an example given in a previous module. We will walk through how this script
captures some of the elements discussed in this module.

[Resources for exercise]

[Video of exercise]

---------------------------------------------------------------------------

> "Languages have
*syntax*, that is, rules that define what is grammatically legal and what is not. 
Programming languages are picky on the grammatical side: you have to say it right or there
will be a complaint." [@kernighan2011d]



> "One way to write R code is simply to enter it interactively at the command line... This 
interactivity is beneficial for experimenting with R or for exploring a data set in a casual
manner. ... However, interactively typing code at the R command line is a very bad approach from 
the perspective of recording and documenting code because the code is lost when R is shut down. 
A superior approach in general is to write R code in a file and get R to read the code from the file." [@murrell2009introduction]

> "The good news is that a computer is a general-purpose machine, capable of performing 
any computation. Although it only has a few kinds of instructions to work with, it can 
do them blazingly fast, and it can largely control its own operation. The bad news is 
that it doesn't do anything itself unless someone tells it what to do, in excruciating
detail. A computer is the ultimate sorcere's apprentice, able to follow instructions
tirelessly and without error, but requiring painstaking accuracy in the 
specification of what to do." [@kernighan2011d]

> "The features of R are organized into separate bundles called *packages*. The standard R 
installation includes about 25 of those packages, but many more can be downloaded from CRAN and
installed to expand the things that R can do. ... Once a package is installed, it must be 
*loaded* within an R session to make the extra features available. ... Of the 25 packages
that are installed by default, nine packages are *loaded* by default when we start a new
R session; these provide the basic functionality of R. All other packages must be loaded
before the relevant features can be used." [@murrell2009introduction]

> "The R environment is the software used to run R code." [@murrell2009introduction]

> "Programming in R is carried out, primarily, by manipulating and modifying data structures.
These different transformations are carried out using functions and operators. In R, 
virtually every operation is a function call, and though we separate our discussion into
operators and function calls, the distinction is not strong ... The R evaluator and
many functions are written in C but most R functions are written in R itself." 
[@gentleman2008r]
