---
title: "Protocol: Estimating bacterial loads from plating samples at different dilutions"
date: "3/1/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This protocol provides details on how you can estimate the bacterial load in
samples based on data from plating those samples at different dilutions. The
question being answered through this protocol of data analysis is: How much
viable (i.e., replicating) bacteria are in each of your samples?

You can find out by **plating** the sample at different **dilutions** and
counting the **colony-forming units (CFUs)** that are cultured on each plate.
You put a sample on a plate with a medium they can grow on and then give them
time to grow. The idea is that individual bacteria from the original sample end
up randomly around the surface of the plate, and any that are viable (able to
reproduce) will form a new colony that, after a while, you'll be able to see.

To count the number of colonies, you need **a "just right" dilution** (likely
won't know what this is until after plating) to have a **countable plate**. If
you have **too high** of a dilution (i.e., one with very few viable bacteria),
randomness will play a big role in the CFU count, and you'll estimate the
original with more variability. If you have **too low** of a dilution (i.e., one
with lots of viable bacteria), it will be difficult to identify separate
colonies, and they may complete for resources. (The pattern you see when the
dilution is too low (i.e., too concentrated with bacteria) is called a
*lawn*---colonies merge together). To translate from diluted concentration to
original concentration, you can then do a back-calculation, incorporating both
the number of colonies counted at that dilution and how dilute the sample was.

## Laboratory description

These data result from plating each sample at several different dilutions and
counting the colony-forming units (CFUs) that are cultured on each plate. Each
sample was on a plate with a medium they can grow on and then given [time] to
grow. On each plate, the individual bacteria from the original sample should
have ended up spread randomly around the surface of the plate, and any that are
viable (able to reproduce) formed a colony that, by the end of the procedure,
can be counted.

[Image of plating]

Each sample was taken from a single mouse from the experiment. These mice came
from different experimental groups, with several replicates (mice) per 
experimental group. The experimental groups in this case were: 

[table of experimental groups]

## Data description

Data are collected for this process into a spreadsheet that is set up in a
"tidy" format used by our laboratory to collect this type of data. Here are the
first few rows of the data:

```{r echo = FALSE}
library(tidyverse)
cfu_data <- read_csv("cfu_data.csv")
head(cfu_data)
```

Each row represents the number of bacterial colonies counted after plating a
certain sample at a certain dilution. Columns are included with values for the
experimental group of the sample (`group`), the specific ID of the sample within
that experimental group (`replicate`, e.g., `2-A` is mouse A in experimental
group 2), the dilution level for that plating (`dilution`), and the number of
bacterial colonies counted in that sample (`CFUs`).

## Reading data into R

The data are stored in a comma-separated plain text file called "cfu_data.csv". 
They can be read into R using the following code: 

```{r}
library(tidyverse)
cfu_data <- read_csv("cfu_data.csv")
head(cfu_data)
```

You will need to be sure that the data file ("cfu_data.csv" in this case) is
in the working directory of your R session. You can check the current working
directory in R with the call `getwd()`, and you can list all files in the 
current working directory with the call `list.files()`. These function calls
can be useful to check to make sure that this file is in your working directory
and, if not, to identify which directory on your computer file system you 
should move it to.

Once you run this command, the data will be available in your R session in the
object `cfu_data`. You can see the first few rows by running: 

```{r}
head(cfu_data)
```

You can get a summary of the data by running: 

```{r}
summary(cfu_data)
```

## Exploring the data and quality checks

It is helpful to explore the data once you read it in, to check for quality
control issues and other characteristics of the data from a particular
experiment. This section provides code examples for this quality control.

First, it can be helpful to ensure that the data includes the number of
experimental groups, and the number of replicates within each group, that
you are expecting. First, we can determine that there are eight experimental 
groups, with between eight and twelve samples (replicates) in each group: 

```{r}
cfu_data %>% 
  group_by(group) %>% 
  count()
```

It is also helpful to see the distribution of CFUs at each dilution level, which 
can be plotted with the following code: 

```{r out.width = "0.8\\textwidth", fig.align = "center"}
cfu_data %>% 
  ggplot(aes(x = CFUs)) + 
  geom_histogram() + 
  facet_wrap(~ dilution)
```
In this case, most bacterial counts are zero at all dilution levels. However, 
there are more non-zero CFUs at dilution level 0 and more as you move to higher
dilution levels. The highest CFU count at any dilution level is a little over
50, for dilution level 1. Most non-zero CFU counts are at or below 30, 
regardless of dilution level.

## Identifying a good dilution for each sample

## Estimating bacterial load for each sample

## Outputing final estimates